Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Pass through ARG section

Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given locus, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

The ``succinct tree sequence'' (STS) is an ARG data structure
that is both general (in terms of the types of ancestry that can be described)
and computationally efficient \cite{Wong2024}.
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
STS has been extended and applied to
forward-time simulations \cite{Kelleher2018efficient,Haller2019},
calculation of population genetics statistics \cite{Ralph2020} and
ARG inference \cite{Kelleher2019,Wohns2022}.
The STS is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations.
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
The genetic inheritance between a pair of nodes along a segment of genome
is defined by the edge ($l$, $r$, $p$, $c$),
which states that child node $c$ inherited its genome from parent node $p$
from left coordinate $l$ to right coordinate $r$.
A site defines a position on the genome and the ancestral state (allele) at that site.
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

Given the node, edge, site and mutation tables,
we can efficiently construct the local genealogical trees along the genome (arising from recombination)
and perform a range of calculations efficiently
by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}.
These algorithms have led to performance increases of several orders of magnitude over
the state-of-the-art in a range of applications
\cite{Kelleher2016,Kelleher2018efficient,Kelleher2019,Ralph2020,Baumdicker2022}.
The STS encoding is also very concise, allowing, for example,
for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019},
and so it is particularly well suited to representing large SARS-CoV-2 genealogies.

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
Tskit is widely used in population genetics, and has been incorporated into many downstream applications
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2023,Mahmoudi2022,Petr2023,Rasmussen2023,Zhang2023}).

>>>>

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024}.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.

% TODO merge this in
Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{Grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given locus, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
The topology of an Ancestral Recombination Graph (ARG) can be defined 
% genome seems better than "haplotype" here, and will be clear what it means
% in the viral context?
as a set of \emph{nodes}, representing a genome that existed at some specific time,
and \emph{edges} that describe the genetic inheritance relationships between those 
genomes~\cite{Wong2024}. 
Each edge defines the parent and child nodes and a genomic interval 
over which inheritance occurs. The ``succinct tree sequence'' ARG encoding defines 
these relationships in a simple and concise tabular format, 
sufficient to represent arbitrarily complex patterns of
recombination~\cite{Wong2024} and providing the basis for a range of highly
efficient data processing
algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
The mature and feature-rich tskit library has interfaces in Python, C and Rust,
extensive documentation,
and is the basis of a growing software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely}
tools.

Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format used 
by UShER, tskit incorporates mutational information directly in the data 
model. This is done using the site (defining the genome position and 
ancestral state) and mutation (defining the site, node, and derived state)
tables. In addition, tskit provides a flexible system for attaching metadata
to all aspects of the data model, which can either be in JSON format 
for flexibility (used heavily by sc2ts for debugging information) 
or in a vectorisable packed binary format for efficiency (used to attach
sample identifiers and pango assignments in the final ARGs). 
Tskit also provides a ``provenance'' table, which provides a mechanism
for recording software version and parameter information through each
step of complex processing pipelines.
